---
title: "SingleR - multiple reference"
author: "Layaa Sivakumar"
date: "2025-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```

# Exploring SingleR: Automatic annotation for single-cell RNA-seq data using multiple reference datasets

Using multiple reference sets has many advantages:

* The diversity in the available cell type labels increases
* Results of annotation may be more accurate 
* Allows cross-validation of annotation i.e., if a cell is labelled as monocyte in multiple references, it is more likely to actually be a monocyte. 


```{r}
library(SingleR)
library(celldex)
library(Seurat)
library(tidyverse)
library(pheatmap)

```

The dataset I'll be working on for this project is single cell RNA-seq data derived from human peripheral blood mononuclear cells (PBMCs) of a healthy female donor aged 25-30. The data was obtained by 10x Genomics from AllCells. 

```{r, Load data}
# Load data 

hdf5_obj <- Read10X_h5(filename = "C:\\Users\\layaa\\Documents\\2024\\Homework\\PD-project-datasets\\singleR\\20k_PBMC_3p_HT_nextgem_Chromium_X_filtered_feature_bc_matrix.h5",
                       use.names = TRUE,
                       unique.features = TRUE)

pbmc.seurat <- CreateSeuratObject(counts=hdf5_obj)

```


```{r, Filtering and QC}
# Filtering and QC

# These parameters filter out cells with few genes detected and high mitochondrial genes, and genes that are expressed at low levels across all cells
pbmc.seurat$mitoPercent <- PercentageFeatureSet(pbmc.seurat, pattern = '^MT-')
pbmc.seurat.filtered <- subset(pbmc.seurat, subset = nCount_RNA > 800 &
         nFeature_RNA > 500 &
         mitoPercent < 10)

```


```{r Standard sc workflow}
# Standard workflow
pbmc.seurat.filtered <- NormalizeData(object = pbmc.seurat.filtered)
pbmc.seurat.filtered <- FindVariableFeatures(object = pbmc.seurat.filtered)
pbmc.seurat.filtered <- ScaleData(object = pbmc.seurat.filtered)
pbmc.seurat.filtered <- RunPCA(object = pbmc.seurat.filtered)
pbmc.seurat.filtered <- FindNeighbors(object = pbmc.seurat.filtered, dims = 1:20)
pbmc.seurat.filtered <- FindClusters(object = pbmc.seurat.filtered)
pbmc.seurat.filtered <- RunUMAP(object = pbmc.seurat.filtered, dims = 1:20)

# running steps above to get clusters
View(pbmc.seurat.filtered@meta.data)
DimPlot(pbmc.seurat.filtered, reduction = 'umap')
```

For this dataset, I will be using 2 references from the celldex package: the human primary cell atlas data and the database immune cell expression data. 

```{r}
# for pbmc data, we will use two datasets
hpca <- celldex::HumanPrimaryCellAtlasData()
dice <- celldex::DatabaseImmuneCellExpressionData()

```

There are different strategies we can use to annotate cells using multiple references with singleR. We will be looking at all 3 of them. 

The first uses reference-specific labels, where we include a prefix to the label that indicates which reference the annotation comes from. Then, we combine all the references into one dataset that includes only the shared genes. 

```{r}
# ...1. Strategy 1: Using reference-specific labels ----------
# adding ref info to labels
hpca$label.main <- paste0('HPCA.', hpca$label.main)
dice$label.main <- paste0('DICE.', dice$label.main)

# create a combined ref based on shared genes 
shared <- intersect(rownames(hpca), rownames(dice))
combined <- cbind(hpca[shared,], dice[shared,])

# run singleR using combined ref
# saving counts into a separate object
pbmc_counts <- GetAssayData(pbmc.seurat.filtered, slot = 'counts')

comb.res1 <- SingleR(test = pbmc_counts, ref = combined, labels = combined$label.main)
table(comb.res1$labels)

pbmc.seurat.filtered$com.res1.labels <- comb.res1[match(rownames(pbmc.seurat.filtered@meta.data), rownames(comb.res1)), 'labels']
View(pbmc.seurat.filtered@meta.data)

DimPlot(pbmc.seurat.filtered, reduction = 'umap', group.by = 'com.res1.labels', label = TRUE)

```

The second strategy does not process the references before passing them into singleR. Instead, both references are passed in a list. After the annotation is complete, we can extract which reference dataset the final label came from, and the differentially expressed genes that define each label per reference. 

```{r}
# ...2. Strategy 2: Comparing scores across references ----------
# remove prefixes from labels
hpca$label.main <- gsub('HPCA\\.','', hpca$label.main)
dice$label.main <- gsub('DICE\\.','', dice$label.main)

comb.res2 <- SingleR(test = pbmc_counts, 
        ref = list(HPCA = hpca, DICE = dice),
        labels = list(hpca$label.main, dice$label.main))

# Check the final label from the combined assignment.
table(comb.res2$labels)

# which reference scored best for which label?
grouping <- paste0(comb.res2$labels,'.', comb.res2$reference)
best_ref <- as.data.frame(split(comb.res2, grouping))

# get de. genes from each individual references
metadata(comb.res2$orig.results$HPCA)$de.genes
metadata(comb.res2$orig.results$DICE)$de.genes

# Combined diagnostics
plotScoreHeatmap(comb.res2)

# UMAP
pbmc.seurat.filtered$comb.res2.labels <- comb.res2[match(rownames(pbmc.seurat.filtered@meta.data), rownames(comb.res2)), 'labels']
DimPlot(pbmc.seurat.filtered, reduction = 'umap', group.by = 'comb.res2.labels', label = TRUE)

```

The third strategy involves using the ontology term held in the label.ont column in the reference datasets as the annotation label. This standardizes labels between references so that there 

```{r}
# ...3. Strategy 3: Using Harmonized Labels ----------

hpca.ont <- celldex::HumanPrimaryCellAtlasData(cell.ont = 'nonna')
dice.ont <- celldex::DatabaseImmuneCellExpressionData(cell.ont = 'nonna')

# Using the same sets of genes:
shared <- intersect(rownames(hpca.ont), rownames(dice.ont))
hpca.ont <- hpca.ont[shared,]
dice.ont <- dice.ont[shared,]

# Showing the top 10 most frequent terms:
tail(sort(table(hpca.ont$label.ont)),10)
tail(sort(table(dice.ont$label.ont)), 10)

# using label.ont instead on label.main while running SingleR

com.res3 <- SingleR(test = pbmc_counts,
        ref = list(HPCA = hpca.ont, DICE = dice.ont),
        labels = list(hpca.ont$label.ont, dice.ont$label.ont))


table(com.res3$labels)

pbmc.seurat.filtered$com.res3.labels <- com.res3[match(rownames(pbmc.seurat.filtered@meta.data), rownames(com.res3)), 'labels']
DimPlot(pbmc.seurat.filtered, reduction = 'umap', group.by = 'com.res3.labels', label = TRUE)

```

The ontology terms can then be mapped back to the labels using the built in mapping TSV file in the celldex package. The process is shown here: 

```{r}
# To map cell ontology terms
# get the mapping tsv files 

colData(hpca.ont)
colData(dice.ont)

hpca.fle <- system.file("mapping","hpca.tsv", package = "celldex")
hpca.mapping <- read.delim(hpca.fle, header = F)

dice.fle <- system.file("mapping","dice.tsv", package = "celldex")
dice.mapping <- read.delim(dice.fle, header = F)

```


```{r}
# map the assigned ontology terms to the label for each cell 

hpca_set <- trial %>%
  mutate(com.res3.main.labels = hpca.mapping$hpca.label.res3[match(com.res3.labels, hpca.mapping$com.res3.labels)])

dice_set <- hpca_set %>%
  filter(is.na(com.res3.main.labels)) %>%
  mutate(com.res3.main.labels = dice.mapping$V1[match(com.res3.labels, dice.mapping$V2)])

ont.mapping <- hpca_set %>% 
  filter(!is.na(com.res3.main.labels)) %>%
  rbind(., dice_set)

pbmc.seurat.filtered$com.res3.ont.mapping <- ont.mapping[match(rownames(pbmc.seurat.filtered@meta.data), rownames(ont.mapping)), "com.res3.main.labels"]

DimPlot(pbmc.seurat.filtered, reduction = 'umap', group.by = 'com.res3.ont.mapping', label = TRUE)
```






